The Unicode Cookbook \linebreak for Linguists

### 2015-11-13 book/chapters/ipa_background.tex
splitting and readability revisions

% ==========================
\chapter{The International Phonetic Alphabet (IPA)}
\label{the-international-phonetic-alphabet}
% ==========================

@@

Established in 1886, the International Phonetic Association (henceforth
Association) has long maintained a standard alphabet, the International Phonetic
Alphabet (IPA), which is a common standard in linguistics to transcribe sounds
of spoken languages. It was first published in 1888 as an international system
of phonetic transcription for oral languages and for pedagogical purposes. It
contained phonetic values for English, French and German. Diacritics for length
and nasalization were already present in this version, and the same symbols are
still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA alphabet was a list of symbols with pronunciation examples
from words in different languages. In 1900 the symbols were first organized into
chart and were given phonetic feature labels for consonants (for manner of
articulation e.g.\ `plosives', `nasales', `fricatives'; for place of
articulation, e.g.\ `bronchiales', `laryngales', `labiales') and for vowels
(e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout the last
century, the structure of the chart has changed with increases in phonetic
knowledge, and thus, like notational systems in other scientific disciplines,
the IPA reflects facts and theories of phonetic knowledge that have developed
over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, this is part of
the Association's mandate. These changes are captured in the revisions to the
IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there has been several revisions, many minor. For example,
articulation labels -- what often are call features even though the IPA avoids
this term -- have changed (e.g.\ terms like `lips', `throat' or `rolled' are not
used anymore). Phonetic symbol values have changed (e.g.\ voiceless is no longer
marked by <h>). Symbols have been dropped (e.g.\ the caret diacritic denoting
`long and narrow' is no longer used). And many symbols have been added to reflect
contrastive sounds found in the world's very diverse phonological systems.

Although IPA began its life as a pedagogical tool, from its earliest days the 
Association aimed to provide ``a separate sign for each distinctive sound; 
that is, for each sound which, being used instead of another, in the same 
language, can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive 
sounds became later known as \textsc{phonemes} and the IPA has developed historically 
into a notational devise with a strictly segmented phonemic view. A phoneme is an 
abstract theoretical notion derived from an acoustic signal as produced by speakers 
in the real world. Therefore the IPA contains a number of theoretical assumptions 
about speech and how to analyze speech in written form.

>>>>

% ==========================
\chapter{The International Phonetic Alphabet}
\label{the-international-phonetic-alphabet}
% ==========================

@@

Established in 1886, the International Phonetic Association (henceforth
Association) has long maintained a standard alphabet, the International Phonetic
Alphabet (IPA), which is a common standard in linguistics to transcribe sounds
of spoken languages. It was first published in 1888 as an international system
of phonetic transcription for oral languages and for pedagogical purposes. It
contained phonetic values for English, French and German. Diacritics for length
and nasalization were already present in this first version, and the same
symbols are still used today. 
%\footnote{Also referred to as API, for \textit{Association Phonétique Internationale}.}

Originally, the IPA alphabet was a list of symbols with pronunciation examples
from words in different languages. In 1900 the symbols were first organized into
a chart and were given phonetic feature labels for consonants (e.g.\ for manner
of articulation among others `plosives', `nasales', `fricatives' and for place
of articulation among others `bronchiales', `laryngales', `labiales') and for
vowels (e.g.\ `fermées', `mi-fermées', `mi-ouvertes', `ouvertes'). Throughout
the last century, the structure of the chart has changed with increases in
phonetic knowledge. Thus, similar to notational systems in other scientific
disciplines, the IPA reflects facts and theories of phonetic knowledge that have
developed over time. It is natural then that the IPA is modified occasionally to
accommodate scientific innovations and discoveries. In fact, this is part of the
Association's mandate. These changes are captured in the recurring revisions to
the IPA.
%\footnote{For a detailed history see:
%\url{https://en.wikipedia.org/wiki/History\_of\_the\_International\_Phonetic_Alphabet}}

Over the years there have been several revisions, many minor. For example, the
articulation labels -- what often are call `features' even though the IPA
deliberately avoids this term -- have changed (e.g.\ terms like `lips', `throat'
or `rolled' are not used anymore). Phonetic symbol values have changed (e.g.\
voiceless is no longer marked by <h>). Symbols have been dropped (e.g.\ the
caret diacritic denoting `long and narrow' is no longer used). And many symbols
have been added to reflect contrastive sounds found in the world's very diverse
phonological systems.

Although the IPA began its life as a pedagogical tool, from its earliest days the 
Association aimed to provide ``a separate sign for each distinctive sound; 
that is, for each sound which, being used instead of another, in the same 
language, can change the meaning of a word'' \citep[27]{IPA1999}. Distinctive 
sounds became later known as \textsc{phonemes} and the IPA has developed historically 
into a notational devise with a strictly segmented phonemic view. A phoneme is an 
abstract theoretical notion derived from an acoustic signal as produced by speakers 
in the real world. Therefore the IPA contains a number of theoretical assumptions 
about speech and how to analyze speech in written form.

### 2015-11-13 book/chapters/ipa_meets_unicode.tex
splitting and readability revisions

\begin{comment}
\subsection{IPA (practical) extensions}

The IPA extensions cover speech sounds beyond the sound systems of the world's
languages, including symbols for pathological speech and paralinguistic
functions (e.g.\ Braille?).

- pathological speech
- paralinguistic functions (Braille?)

As noted in Section \ref{}, extensions to the IPA -- in particular
``large-scale'' extensions --

% Englebretson2009.pdf
http://www.ruf.rice.edu/~reng/englebretson2009.pdf

X-SAMPA was an ASCII transliteration and is still purposeful for several
programs

There is also the extensions of the IPA designed for disordered speech, although
the symbols are occasionally used for transcribing phonetic detail in ``normal''
speech or also cross-linguistic
rarities.\footnote{\url{http://phoible.github.io/conventions/}}

% Unicode has its own history; for a summary:
\url{http://www.unicode.org/history/summary.html}

https://www.phon.ucl.ac.uk/home/wells/ipa-unicode.htm

Computer codes for phonetic symbols
EslingGaylord1993

\end{comment}

@@

Summarizing the pitfalls as discussed in this chapter, we propose to define 
three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@ These 
three encodings are subsets of each other, i.e.~strict-IPA is more restricted 
than valid-IPA, which it turn is more restricted than widened-IPA.\@ Informally 
speaking, valid-IPA represents the current state of the IPA \citep{IPA2005}. 
Strict-IPA represents a more constrained version of IPA, while widened-IPA is a 
slightly extended version of IPA, allowing a few more symbols.

@@

textbf{NOTE: I HAVE ADDED VARIOUS TONE DIACRITICS TO VALID-IPA THAT ARE NOT IN
THE OFFICIAL IPA SUMMARY. SHOULD THEY BE INCLUDED?}

>>>>

@@

Summarizing the pitfalls as discussed in this chapter, we propose to define
three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@
Informally speaking, valid-IPA represents the current state of the IPA
\citep{IPA2005}. Strict-IPA represents a more constrained version of IPA, while
widened-IPA is a slightly extended version of IPA, allowing a few more symbols.

@@
